Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
capitalize Figure

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages (e.g.
PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g. Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

@@

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} - sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

To also support these properties for irreversible operations that
cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
non-transactional code in general), we use \emph{irrevocable} or
\emph{inevitable transactions}~\cite{blundell06,spear08}. These transactions are always
guaranteed to commit, which is why they always have to win in case
there is a conflict with another, normal transaction. There is always
at most one such transaction running in the system, thus their
execution is serialised. With this guarantee, providing \emph{strong
isolation} and \emph{serialisability} between non-transactional code
is possible by making the current transaction inevitable right before
running irreversible operations.

@@

In this section, we will present the general idea of how the TM model
is implemented. Especially the aspects of providing isolation and
atomicity, as well as conflict detection are explained. We try to do
this without going into too much detail about the implementation.  The
later section \ref{sub:Low-level-Implementation} will discuss it in
more depth.

\subsubsection{Memory Segmentation}

A naive approach to providing complete isolation between threads is to
partition the virtual memory of a process into $N$ segments, one per
thread. Each segment then holds a copy of all the memory available to
the program. Thus, each thread automatically has a private copy of
every object that it can modify in complete isolation from other
threads.

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads~/ segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in figure \ref{fig:Segment-Addressing}.

x86-CPUs provide a feature called \emph{memory segmentation}. It
performs this translation from a $SO$ to a LA directly in hardware.  We
can use the segment register $\%gs$, which is mostly unused in current
applications. When this register points to a thread's segment start
address, we can instruct the CPU to perform the above translation from
a reference of the form $\%gs{::}SO$ to the right LA on its own. This
process is efficient enough that we can do it on every access to an
object.

In summary, we can use a single $SO$ to reference the same object in all
threads, and it will be translated by the CPU to a LA that always
points to the thread's private version of this object. Thereby,
threads are fully isolated from each other. However, $N$ segments
require $N$-times the memory and modifications on an object need to be
propagated to all segments.

@@

\subsubsection{Page Sharing}

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages.

The mapping generated by \lstinline!mmap()! is initially linear but
can be changed arbitrarily. Especially, we can remap so that multiple
virtual memory pages map to a single virtual file page. This is what
we use to share memory between the segments, since we then only
require one page of physical memory for all of them.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.8]{\string"mmap pages\string".pdf}
  \caption{\texttt{mmap()} Page Mapping\label{fig:mmap()-Page-Mapping}}
\end{figure}

As illustrated in figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of
virtual file pages. This is the share-nothing configuration where
all threads have private versions of all objects.

We then designate segment~0 to be the \emph{sharing-segment}. No
thread gets this segment assigned to it, it simply holds the pages
shared between threads. So in step (II), we remap all virtual pages of
the segments~$>0$ to the file pages of our sharing-segment. This is
the fully-shared configuration where no threads have private versions
of any objects.

During runtime, we can then privatise single pages in segments~$>0$
again by remapping single pages as seen in (III). All objects in that
page now have a private version in some thread.

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is cost-free: it is part of the normal operations done by the
Memory Management Unit (MMU) of the CPU on every memory access.

In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
either a physical memory location which is shared between several
threads~/ segments or to a location in memory private to the segment~/
thread. Page sharing makes the memory segmentation model for isolation
memory efficient again.

\subsubsection{Isolation: Copy-On-Write}

We now use these mechanisms to provide isolation for transactions.  We
implement a \emph{Copy-On-Write (COW)} on the level of
pages~\footnote{Conflict detection still occurs on the level of
objects.}. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

More precisely, this is done with a write barrier that detects that we are
about to write to an object that we did not record
in the write-set yet.  When this occurs, the slow-path of the write barrier
will also check if the page (or pages) containing the object is still
shared, and if so, privatise it.  This is done by remapping and copying
the page to the thread's own segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.

\subsubsection{Isolation: Barriers}

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object, except if
we statically know an object to be readable or writable already.
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
  transaction. Since our two-step address translation automatically
  resolves the reference to the private version of the object on every
  access anyway, the read barrier does not have the job to find the
  private version. This job is fully performed by the CPU so that
  read barriers have very little work to do.
\item [{Write~Barrier:}] Adds the object to the read and write set of
  the current transaction and checks if all pages of the object are
  private, doing COW otherwise.\\
  Furthermore, we currently allow only one transaction modifying an
  object at a time. To ensure this, we acquire a write lock on the object
  and also eagerly check for a write-write conflict at this point. If
  there is a conflict, we do some contention management to decide which
  transaction has to wait or abort.\\
  Eagerly detecting this kind of conflict is not inherent to our
  system, future experiments may show that we want to lift this
  restriction.
\end{description}

\subsubsection{Atomicity: Commit \& Abort}

To provide atomicity for a transaction, we want to make changes globally
visible on commit. We also need to be able to completely abort a
transaction without a trace, like it never happened.
\begin{description}
\item [{Commit:}] If a transaction commits, we synchronise all threads
  so that all of them are waiting in a safe point. In the committing
  transaction, we go through all objects in the write set and check if
  another transaction in a different segment read the same object.
  Conflicts are resolved again by either the committing or the other
  transaction waiting or aborting.\\
  We then push all changes of modified objects in private pages to all
  the pages in other segments, including the sharing-segment (segment
  0).
\item [{Abort:}] On abort the transaction will forget about all the
  changes it has done. All objects in the write set are reset by
  copying their previous version from the sharing-segment into the
  private pages of the aborting transaction.\\
  Re-sharing these pages instead of resetting the changes would not
  only be slower because it involves a system call, it is also very
  likely the page would get privatised again immediately after re-trying
  the transaction.  Since this again involves a full page copy,
  resetting should be faster than re-sharing.
\end{description}

\subsubsection{Summary}

We provide isolation between transactions by privatising the pages of
the segments belonging to the threads the transactions run in.  To
detect when and which pages need privatisation, we use write barriers
that trigger a COW of one or several pages. Conflicts, however, are
detected on the level of objects; based on the concept of read and
write sets. Barriers before reading and writing add objects to the
corresponding set; particularly detecting write-write conflicts
eagerly.  On commit, we resolve read-write conflicts and push
modifications to other segments. Aborting transactions simply undo
their changes by copying from the sharing-segment.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g. it shares the
write barrier). The close integration helps in order to know more
about the lifetime of an object, as will be explained in the following
sections.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers}
and already mentioned in section \ref{sub:Setup}.

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section \ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

Overall, we believe that disabling it on all benchmarks except the
performance benchmarks in section \ref{sec:performance-bench} is better
because we can minimise non-determinism. We also do not want to depend
on the capabilities of the JIT in these experiments.

@@

Second, there are areas in memory private to each segment (see
section \ref{sub:Setup}). The Nursery for example is 4~MiB in
size. Additionally, the area used for the read markers is lazily
mapped by the kernel and only needed for objects in the old object
space. Thus, it can be at most the total size of memory used by
old objects divided by 16.

@@

In figure \ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For PyPy-STM the average memory requirement is 21~MiB and there are
$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time. Compared to normal PyPy, we are missing a
memory optimisation to store small objects in a more compact
way. Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g. the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
We expect to improve this aspect in the future; in this paper we
want to focus first on performance.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}

@@

For the results in figure \ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in figure \ref{fig:performance-nojit}.

@@

The results are presented in figure \ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.

>>>>

\conferenceinfo{DLS 2014}{to be supplied}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

Dynamic languages like Python, PHP, Ruby, and JavaScript are usually
regarded as very expressive but also very slow. In recent years, the
introduction of just-in-time compilers (JIT) for these languages
(e.g.\ PyPy~\cite{cfbolz09}, V8~\cite{kevin10}, IonMonkey~\cite{ionmonkey})
started to change this perception by delivering
good performance that enables new applications. However, a parallel
programming model was not part of the design of those languages. Thus,
the reference implementations of e.g.\ Python and Ruby use a single,
global interpreter lock (GIL) to serialise the execution of code in
threads.

@@

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

To also support these properties for irreversible operations that
cannot be undone when we abort a transaction (e.g.\ I/O, syscalls, and
non-transactional code in general), we use \emph{irrevocable} or
\emph{inevitable transactions}~\cite{blundell06,spear08}. These transactions are always
guaranteed to commit, which is why they always have to win in case
there is a conflict with another, normal transaction. There is always
at most one such transaction running in the system, thus their
execution is serialised. With this guarantee, providing \emph{strong
isolation} and \emph{serialisability} between non-transactional code
is possible by making the current transaction inevitable right before
running irreversible operations.

@@

In this section, we will present the general idea of how the TM model is
implemented.  The later section~\ref{sub:Low-level-Implementation} will
discuss it with more technical details.

\subsubsection{Memory Segmentation and Page Sharing}

A naive approach to providing complete isolation between threads is to
partition the virtual memory of a process into $N$ segments, one per
thread.  Each segment then holds a complete copy of the memory logically
seen by the program.  All threads read and write data inside their own
segment; exchanges between segments occur only at commit-time.  So far,
this is a model that would also work for distributed transactional
memory.

Obviously, if we are in a shared-memory setting, this model is extremely
wasteful: at a given time, most objects are identical in all, or most,
segments.  They differ only in segments that have modified but not yet
committed the object.  So we use a trick of the Memory Management Unit
(MMU) of the CPU.  At the coarse unit of one page of memory (4096
bytes), we use remappings so that, in general, pages within segments are
\emph{shared} with each other if they would contain a fully identical
copy.  Conversely, when one segment needs to write to one object
contained in a page, the complete page is first \emph{privatised}: it is
duplicated and the copy is remapped so that it will be visible in this
segment (only).  This process is illustrated in
figure~\ref{fig:Page-Remapping}.

The result that we get is that the virtual address space of the process
is bloated -- by a factor $N$ -- but the real memory usage is not --
because most pages are shared with their copies in most other segments.

\subsubsection{Memory Segmentation on the x86}

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is precisely called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment register
$\%gs$, which is mostly unused in current applications, to point to the
thread's segment start address.\footnote{The other segment register
$\%fs$ is typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for a
similar purpose: thread-local data -- but a lot of it.}  The translation
from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
before any CPU instruction that reads or writes memory using a $SO$
address.  This is very efficient and some compilers support it natively
(e.g.\ clang).

@@

\subsubsection{Conflict detection}

What we described so far has no mesurable performance impact by itself,
but is missing conflict detection to be part of a full-fledged STM
system.  This requires adding \emph{barriers} to the program to register
the objects into the read or write set, and adapting the commit
protocol in consequence.

\begin{description}

\item [{Read~Barrier:}] A major contribution to our performance results
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  We managed to
  design the read barrier so that it is done by one memory write
  (setting a flag) into non-shared memory.  This is done with a
  segment-local array of flags indexed by a number derived from the $SO$
  of the object.  Unlike other STM systems, the read barrier does not
  have to find the private version of some object -- our two-step
  address translation automatically resolves the reference to the
  private version on every access anyway.  This means the read barrier
  is not constrained to execute before the actual read -- both the
  compiler and the CPU are free to reorder them.

\item [{Write~Barrier:}] This is triggered by checking a flag in the
  header of the object.  When the flag is set, we call the slow path,
  which adds the object into the write set and unconditionally resets
  the flag.  We eagerly detect write-write conflicts by allowing only
  one transaction modifying an object at a time (by acquiring a write
  lock).\footnote{Eager write-write detection is not inherent to our
  approach; future experiments may show that we want to lift this
  restriction.}  This slow path is also responsible for privatising
  pages,\footnote{The write set contains objects, not pages.} as
  described above.  Interestingly, we are using a generational garbage
  collector (not described here) which requires a similar write barrier
  anyway -- so the same header flag is used for both cases.  The flag
  means ``something needs to be done before writing to this object'',
  and the slow path logic determines what it is.

\item [{Commit:}] When we want to commit a transaction, we need to check
  for read-write conflicts.  Such conflicts cannot be detected in the
  minimal read barrier, but can be detected during commit: the case to
  look for is that the committing transaction wrote to an object that
  another transaction has already read.\footnote{Which is why the read
  sets can be just sparsely populated arrays of flags: we never walk the
  whole read set.} A priori, checking other transactions' read sets
  would involve reading concurrently modified memory: a different thread
  could be issuing a read barrier in parallel with an unprotected
  commit.  To prevent this from occurring, we must suspend all other
  threads at known safe points.  This is a trade-off for the simplicity
  of our read barriers -- which is well worth it in our case, due to the
  very high number of read barriers when compared to the number of
  commits and the (low) total number of threads.

\end{description}

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not includes all the new objects created by the current transaction
(there is typically a very high turn-around rate in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

Privatised pages are re-shared during major collection cycles.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g.\ it shares the
write barrier). The close integration helps in order to know more
about the lifetime of an object, as will be explained in the following
sections.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

To add the object to the read set, for us it is enough to mark it as
read. Since this information needs to be local to the segment, we need
to store it in private pages. The area is called \emph{read markers}
and already mentioned in section~\ref{sub:Setup}.

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see~\ref{subsub:contentionmanagement}).

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

Overall, we believe that disabling it on all benchmarks except the
performance benchmarks in section~\ref{sec:performance-bench} is better
because we can minimise non-determinism. We also do not want to depend
on the capabilities of the JIT in these experiments.

@@

Second, there are areas in memory private to each segment (see
section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
size. Additionally, the area used for the read markers is lazily
mapped by the kernel and only needed for objects in the old object
space. Thus, it can be at most the total size of memory used by
old objects divided by 16.

@@

In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For PyPy-STM the average memory requirement is 21~MiB and there are
$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time. Compared to normal PyPy, we are missing a
memory optimisation to store small objects in a more compact
way. Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g.\ the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
We expect to improve this aspect in the future; in this paper we
want to focus first on performance.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}

@@

For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in Figure~\ref{fig:performance-nojit}.

@@

The results are presented in Figure~\ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.
